Improving the portuguese hyphenation rules

### 2023-07-13 article/article.tex
fix #18 also fix Oxford commas

Another important matter to consider is ambiguities that might be created when
a word is partitioned during the hyphenation process. In English we should
avoid hyphenations such as \emph{re-cover}, \emph{re-form}, \emph{re-sign}, \emph{the-rapist},
\emph{depart-mental} and \emph{mans-laughter} (in Portuguese, some examples are:
\emph{de-putada}, \emph{fede-ração}, \emph{acu-mula}, \emph{após-tolo},
\emph{cú-bico}). Hyphenations that might lead the reader to pronounce a word
incorrectly should also be prevented. That is the case of
\emph{considera-tion}, in Enslish (and \emph{pe-rigo} in Portuguese).

@@

The general rule for Portuguese hyphenation is to split a word into its
syllables. A syllable is made of a mandatory nucleus, filled by a vowel, and
optional peripheral consonants (before or after the nucleus). In some
situations, the syllabic division does not respect the ethnologic constituents.
The usage of the prefixes \emph{bis-} and \emph{in-} are examples of this
circumstance. The correct syllabifications are \emph{bi-sa-vô},
\emph{i-na-ti-vo} and \emph{i-nobs-tan-te}\footnote{The rule of syllable
division could lead to two possible partitions: \emph{i-nobs-tan-te} and
\emph{in-obs-tan-te}, but the first is preferable.}, where the prefixes are
split into two syllables. But, as pointed previously, it is preferable to keep
morphemes together rather than splitting them apart, therefore the we should
favor the hyphenation \emph{bisa-vô} over \emph{bi-savô} and \emph{ina-tivo}
over \emph{i-nativo}. This last example could also lead to a misunderstanding
since the word \emph{nativo} (\emph{native} in English) emerges from this word
break.

@@

Portuguese uses an alphabetical writing system, ensuing that its spelling is
guided by phonological principles \cite{cagliari2015}. The fact that spelling
reflects, to some extent, the way words are pronounced, means that the
hyphenation of words is also influenced by the phonemic system, where words are
divided into syllables. It is noteworthy that different languages divide words
based on different principles, for example, English tends to be guided by
morphological principles, such as \emph{walk-ing}, \emph{un-happy},
\emph{work-s} and \emph{ear-ly}. Also, other factors affect word hyphenation in
English, such as the presence of doubled consonants and digraphs\footnote{One
should not confuse consonant clusters with digraphs, since the latter represent
only one sound, while the former represent two or more sounds.}. There are also
a myriad of exceptions that might be grasped into rules.
% https://english.stackexchange.com/questions/385/what-are-the-rules-for-splitting-words-at-the-end-of-a-line

Merely stating that an orthographic system is guided by phonological issues
does not necessarily mean that its hyphenation rules directly reflect the way a
word is pronounced -- this is because Portuguese does not have a direct
letter-sound correspondence, and vice versa.  The orthographic system operates
according to its own specific rules. For instance, although vowel clusters that
produce a single sound (i.e., digraphs) are pronounced within a single syllable
in words like \emph{chave}, \emph{ilha}, \emph{shampoo}, \emph{carro}, and
\emph{pássaro}, their representation in writing determines how they are
separated.  Specifically, different consonants within a digraph must remain
together, while different consonants are separated. As a result, we have
\emph{i-lha} but \emph{car-ro}.

@@

\begin{description} 
    \item[Morphological determination:] Prefixes such as \emph{re-},
	\emph{sub-}, \emph{cyber-} and \emph{auto-} require separating the
	prefix from its stem, what runs into phonological issues. For example,
	\emph{reiniciar} and \emph{sublinhar}, as they contain prefixes, should
	be hyphenated as \emph{re-i-ni-ci-ar} and \emph{sub-li-nhar},
	respecting the morphological formation of the word. However,
	considering that the Portuguese language hyphenates its words from
	syllabic phonological correlates, words that require morphological
	information, such as those, do not have their hyphenation performed
	properly. They add up, in our corpus, to twelve words (eight words with
	\emph{re-} and one case with \emph{auto-}, \emph{meta-}, \emph{sub-},
	\emph{cyber-}, \emph{geo-} or \emph{neuro-}).

    \item[Foreignness:] There is a group of 20 words in the corpus that are
	terminologies or words that were incorporated into the Portuguese
	language without full phonological adaptation, such as
	\emph{darwinismo}, \emph{quilowatt} and \emph{esfiha}. The lack of
	adaptation makes the phonological pattern very specific to the word,
	making it impossible to incorporate in \TeX{}'s rules alongside the
	other rules. The solution is to add them to the exception word list.

    \item[Word-initial consonant clusters:] Portuguese has few cases of
	consonant clusters at the beginning of a word. They, in general, are
	etymological remnants and are currently unproductive in the language,
	since there are no neologisms with this pattern. Encounters like
	\emph{ps-} and \emph{pn-} are more frequent, as they are present in
	words like \emph{psicologia} and \emph{pneu}, which have an average
	frequency in the language. They might be predicted by rules, which
	comprise XXX words. However, there are consonant clusters that are
	found in very specific and low frequency words. Although possible, it
	is not worth adding very specific rules using the consonant clusters
	contained in the words \emph{dzeta}, \emph{gnu}, \emph{cnidário},
	\emph{ftálico} and \emph{gnaisse} -- that is, just five words.

    \item[Abbreviations, Acronyms, or Initialisms:] Whether for efficiency, 
	convenience, clarity or jargon, it is usual the usage of a shortening 
	version of a word or a phrase. 
	Abbreviation a way to shorten a word of phrase. Some examples in our
	Portuguese corpus are: 
	\emph{etc.}\footnote{Latin expression \emph{et cetera}, meaning 
	`and other similar things'.}, 
	\emph{Dr.}\footnote{\emph{doutor} (doctor, person with PhD title, but popularly used to designate an erudite individual)}, 
	\emph{Exmo.}\footnote{\emph{Excelentíssimo} (honourable)},
	\emph{cap.}\footnote{\emph{capítulo} (chapter)}, \emph{Univ.}\footnote{Universidade (university)}, 
	\emph{ed.}\footnote{\emph{edição} (edition)}, 
	\emph{s.n.}\footnote{sine nomine, Latin expression meaning `without a name', 
	mostly used in the context of publishing.}.
	Initialism consists of a set of initial 
	letters of words that form an shortened version. They are not always 
	understood by the set of hyphenation rules described in this work, 
	since they do not follow orthographic (or phonotactic) standards of 
	the language. In the corpus, it is possible to find some abbreviations, 
	such as \emph{SESC}\footnote{Serviço Social do Comércio}, 
	\emph{INSS}\footnote{\emph{Instituto Nacional do Seguro Social} (National Institute of Social Security)}, 
	\emph{PCdoB}\footnote{\emph{Partido Comunista do Brasil} (Communist Party of Brazil)},
	\emph{PM}\footnote{\emph{Polícia Militar} (military police)} and 
	\emph{UFRJ}\footnote{Universidade Federal do Rio de Janeiro}.
	Acronym is a type of abbreviation that shortens a phrase by combining
	the first letter (or letters) of each word, creating a new pronounceable word.
	Some example from the corpus are: 
	\emph{Anatel}\footnote{\emph{Agência Nacional de Telecomunicações} (National Telecommunications Agency)},
	\emph{Ovni}\footnote{\emph{Objeto voador não identificado} (unidentified flying object - UFO)},
	\emph{Sida}\footnote{\emph{Síndrome da Imunodeficiência Adquirida} (acquired immunodeficiency syndrome - AIDS)} (in Portugal) and
	\emph{Mercosul}\footnote{\emph{Mercado Comum do Sul} (Southern Common Market).}.
\end{description}

>>>>

Another important matter to consider is ambiguities that might be created when
a word is partitioned during the hyphenation process. In English we should
avoid hyphenations such as \emph{re-cover}, \emph{re-form}, \emph{re-sign}, \emph{the-rapist},
\emph{depart-mental}, and \emph{mans-laughter} (in Portuguese, some examples are:
\emph{de-putada}, \emph{fede-ração}, \emph{acu-mula}, \emph{após-tolo},
\emph{cú-bico}). Hyphenations that might lead the reader to pronounce a word
incorrectly should also be prevented. That is the case of
\emph{considera-tion}, in Enslish (and \emph{pe-rigo} in Portuguese).

@@

The general rule for Portuguese hyphenation is to split a word into its
syllables. A syllable is made of a mandatory nucleus, filled by a vowel, and
optional peripheral consonants (before or after the nucleus). In some
situations, the syllabic division does not respect the ethnologic constituents.
The usage of the prefixes \emph{bis-} and \emph{in-} are examples of this
circumstance. The correct syllabifications are \emph{bi-sa-vô},
\emph{i-na-ti-vo}, and \emph{i-nobs-tan-te}\footnote{The rule of syllable
division could lead to two possible partitions: \emph{i-nobs-tan-te} and
\emph{in-obs-tan-te}, but the first is preferable.}, where the prefixes are
split into two syllables. But, as pointed previously, it is preferable to keep
morphemes together rather than splitting them apart, therefore the we should
favor the hyphenation \emph{bisa-vô} over \emph{bi-savô} and \emph{ina-tivo}
over \emph{i-nativo}. This last example could also lead to a misunderstanding
since the word \emph{nativo} (\emph{native} in English) emerges from this word
break.

@@

Portuguese uses an alphabetical writing system, ensuing that its spelling is
guided by phonological principles \cite{cagliari2015}. The fact that spelling
reflects, to some extent, the way words are pronounced, means that the
hyphenation of words is also influenced by the phonemic system, where words are
divided into syllables. It is noteworthy that different languages divide words
based on different principles, for example, English tends to be guided by
morphological principles, such as \emph{walk-ing}, \emph{un-happy},
\emph{work-s}, and \emph{ear-ly}. Also, other factors affect word hyphenation in
English, such as the presence of doubled consonants and digraphs\footnote{One
should not confuse consonant clusters with digraphs, since the latter represent
only one sound, while the former represent two or more sounds.}. There are also
a myriad of exceptions that might be grasped into rules.
% https://english.stackexchange.com/questions/385/what-are-the-rules-for-splitting-words-at-the-end-of-a-line

Merely stating that an orthographic system is guided by phonological issues
does not necessarily mean that its hyphenation rules directly reflect the way a
word is pronounced -- this is because Portuguese does not have a direct
letter-sound correspondence, and vice versa.  The orthographic system operates
according to its own specific rules. For instance, although consonant clusters that
produce a single sound (i.e., digraphs) are pronounced within a single syllable
in words like \emph{chave}, \emph{ilha}, \emph{shampoo}, \emph{carro}, and
\emph{pássaro}, their representation in writing determines how they are
separated.  Specifically, different consonants within a digraph must remain
together, while different consonants are separated. As a result, we have
\emph{cha-ve}, \emph{i-lha}, and \emph{sham-poo}, but \emph{car-ro}.

@@

\begin{description} 
    \item[Morphological determination:] Prefixes such as \emph{re-},
	\emph{sub-}, \emph{cyber-}, and \emph{auto-} require separating the
	prefix from its stem, what runs into phonological issues. For example,
	\emph{reiniciar} and \emph{sublinhar}, as they contain prefixes, should
	be hyphenated as \emph{re-i-ni-ci-ar} and \emph{sub-li-nhar},
	respecting the morphological formation of the word. However,
	considering that the Portuguese language hyphenates its words from
	syllabic phonological correlates, words that require morphological
	information, such as those, do not have their hyphenation performed
	properly. They add up, in our corpus, to twelve words (eight words with
	\emph{re-} and one case with \emph{auto-}, \emph{meta-}, \emph{sub-},
	\emph{cyber-}, \emph{geo-} or \emph{neuro-}).

    \item[Foreignness:] There is a group of 20 words in the corpus that are
	terminologies or words that were incorporated into the Portuguese
	language without full phonological adaptation, such as
	\emph{darwinismo}, \emph{quilowatt}, and \emph{esfiha}. The lack of
	adaptation makes the phonological pattern very specific to the word,
	making it impossible to incorporate in \TeX{}'s rules alongside the
	other rules. The solution is to add them to the exception word list.

    \item[Word-initial consonant clusters:] Portuguese has few cases of
	consonant clusters at the beginning of a word. They, in general, are
	etymological remnants and are currently unproductive in the language,
	since there are no neologisms with this pattern. Encounters like
	\emph{ps-} and \emph{pn-} are more frequent, as they are present in
	words like \emph{psicologia} and \emph{pneu}, which have an average
	frequency in the language. They might be predicted by rules, which
	comprise XXX words. However, there are consonant clusters that are
	found in very specific and low frequency words. Although possible, it
	is not worth adding very specific rules using the consonant clusters
	contained in the words \emph{dzeta}, \emph{gnu}, \emph{cnidário},
	\emph{ftálico}, and \emph{gnaisse} -- that is, just five words.

    \item[Abbreviations, Acronyms, or Initialisms:] Whether for efficiency, 
	convenience, clarity or jargon, it is usual the usage of a shortening 
	version of a word or a phrase. 
	Abbreviation a way to shorten a word of phrase. Some examples in our
	Portuguese corpus are: 
	\emph{etc.}\footnote{Latin expression \emph{et cetera}, meaning 
	`and other similar things'.}, 
	\emph{Dr.}\footnote{\emph{doutor} (doctor, person with PhD title, but popularly used to designate an erudite individual)}, 
	\emph{Exmo.}\footnote{\emph{Excelentíssimo} (honourable)},
	\emph{cap.}\footnote{\emph{capítulo} (chapter)}, \emph{Univ.}\footnote{Universidade (university)}, 
	\emph{ed.}\footnote{\emph{edição} (edition)}, 
	\emph{s.n.}\footnote{sine nomine, Latin expression meaning `without a name', 
	mostly used in the context of publishing.}.
	Initialism consists of a set of initial 
	letters of words that form an shortened version. They are not always 
	understood by the set of hyphenation rules described in this work, 
	since they do not follow orthographic (or phonotactic) standards of 
	the language. In the corpus, it is possible to find some abbreviations, 
	such as \emph{SESC}\footnote{Serviço Social do Comércio}, 
	\emph{INSS}\footnote{\emph{Instituto Nacional do Seguro Social} (National Institute of Social Security)}, 
	\emph{PCdoB}\footnote{\emph{Partido Comunista do Brasil} (Communist Party of Brazil)},
	\emph{PM}\footnote{\emph{Polícia Militar} (military police)}, and 
	\emph{UFRJ}\footnote{Universidade Federal do Rio de Janeiro}.
	Acronym is a type of abbreviation that shortens a phrase by combining
	the first letter (or letters) of each word, creating a new pronounceable word.
	Some example from the corpus are: 
	\emph{Anatel}\footnote{\emph{Agência Nacional de Telecomunicações} (National Telecommunications Agency)},
	\emph{Ovni}\footnote{\emph{Objeto voador não identificado} (unidentified flying object - UFO)},
	\emph{Sida}\footnote{\emph{Síndrome da Imunodeficiência Adquirida} (acquired immunodeficiency syndrome - AIDS)} (in Portugal) and
	\emph{Mercosul}\footnote{\emph{Mercado Comum do Sul} (Southern Common Market).}.
\end{description}
